\section{\Zoo in practice}
\label{sec:zoo}

\input{figures/zoo}

Before describing the salient features of our language, \Zoo, in \cref{sec:features}, we give an overview of the framework.

\paragraph{From \OCaml to \Zoo.}

First, \OCaml source files are translated into \Zoo by the \texttt{ocaml2zoo} tool.
The \Zoo syntax is given in \cref{fig:zoo}\footnote{More precisely, it is the syntax of the surface language, including many \Coq notations.}, omitting mutually recursive toplevel functions that are treated specifically.
Essentially, \Zoo is an untyped, ML-like, imperative, concurrent programming language.
The supported \OCaml fragment includes: shallow \mintinline{ocaml}{match}, ADTs, records, inline records, atomic record fields, unboxed types, toplevel mutually recursive functions.

% [Gabriel]: it may not be obvious to the reader why Zoo is untyped -- why it is reasonable not to have a type system, despite the known benefits of type systems in clarifying programs and preventing bugs.

For instance, the \mintinline{ocaml}{push} function from \cref{sec:introduction} is translated into:

\begin{minted}{coq}
Definition stack_push : val :=
  rec: "push" "t" "v" =>
    let: "old" := !"t" in
    let: "new_" := "v" :: "old" in
    ifnot: CAS "t".[contents] "old" "new_" then (
      Yield ;;
      "push" "t" "v"
    ).
\end{minted}

\paragraph{Specifications and proofs.}

Second, the user writes specifications for the translated functions and prove them using the \Iris proof mode~\cite{DBLP:journals/pacmpl/KrebbersJ0TKTCD18}.

For instance, the specification for the \mintinline{ocaml}{stack_push} function would be:

\begin{minted}{coq}
Lemma stack_push_spec t ι v :
  <<<
    stack_inv t ι
  | ∀∀ vs, stack_model t vs
  >>>
    stack_push t v @ ↑ι
  <<<
    stack_model t (v :: vs)
  | RET (); True
  >>>.
Proof.
  ...
Qed.
\end{minted}

Here, we use a \emph{logically atomic specification}~\cite{DBLP:conf/ecoop/PintoDG14}, which has been proven~\cite{DBLP:journals/pacmpl/BirkedalDGJST21} to be equivalent to \emph{linearizability}~\cite{DBLP:journals/toplas/HerlihyW90} in sequentially consistent memory models.
